\vspace{-1.3cm} \textmd{\normalsize{High Performance Computing Project}}\\ Parallelization of the Probabilistic Roadmap Method with GPU Acceleration

### 2016-03-29 paper/paper.tex
paper...

The abstract problem is to find a path on a d-dimensional space, which avoids intersecting obstacles. The obstacles are  given by an indicator function
\begin{equation*}
	I: \Omega \rightarrow \{0, 1\}
\end{equation*}
on the domain \( \Omega = [ q_{min,1}, q_{max,1} ] \times ... \times  [ q_{min,d}, q_{max,d} ] \subset \R^d  \) and the goal is to find a continuous way
\begin{equation*}
	\Gamma: [0,1] \rightarrow \Omega
\end{equation*}
between a given start and endpoint, \(\Gamma(0)=q_b,\ \Gamma(1)=q_e \), 
such that
\begin{equation*}
	I(\Gamma(s))=0 \text{\ f.a.\ } s \in [0,1].
\end{equation*}
The probabilistic roadmap (PRM) approach grows a graph \(G=(V,E)\) form \(q_s\) and \(q_e\) by randomly sampling nodes \(v\in V\subset \Omega\).  Two nodes \(v,w \in V\) are connected, if they are near enough each other and the linear connection in between is free from obstacles. The algorithm terminates, when there is a path from \(q_s\) to \(q_e\) on G.
\begin{defi}
For two points \(q_1\) and \(q_2\) we define the connection with stepsize \(h>0\) as
	\begin{equation*}
		[q_1,q_2]_h \defgl \{ q = \lambda q_1+(1-\lambda) q_2\ |\ \lambda \in [0,1], \norm{q-q_1} \in \N_0 h \}.
	\end{equation*}
	We say, \(q_1\) and \(q_2\) are connected (with stepsize \(h>0\)) on the map \(\Omega\), if
	\(I(q)=0\) for all \(q \in [q_1, q_2]_h\).
\end{defi}
\begin{alg1} (Probabilistic Roadmap in general)\\
	Input: \(q_s, q_e\in \Omega, h>0, d_0>0\)	\\
	Output: \(q_s=q_1, ... ,q_n=q_e\in \Omega\), 
		such that all \(q_i\) and \(q_{i+1}\) are connected with stepsize \(h\)
	\begin{itemize}
		\item Initialize graph \(G=(V,E)\) with \(V=\{q_s, q_e\}\), 
			if \(q_s\) and \(q_e\) are connected, \(E \leftarrow\{ q_s, q_e\}\)
		\item While there exists no path in G between \(q_s\) and \(q_e\) do
		\begin{itemize}
			\item Sample \(q\in \Omega\) from a probability density  \(p(\ \cdot\ , G)\), resample until \(I(q)=0\)
			\item \(V \leftarrow{q}\)
			\item For all nodes \(\tilde{q} \in V\) which are near enough (\( \norm{q- \tilde q} \le d_0 \)) and
				connected with \(q\) with stepsize h, 
				\(E \leftarrow \{q, \tilde{q}\}\)
		\end{itemize}
		\item Determine the shortest path \(q_1, ... , q_n\in V\) on \(G\) from \(q_s\) to \(q_e\).
	\end{itemize}
\end{alg1}

@@

%\ \\
\begin{center} 
\begin{tikzpicture}[auto]%node distance=0.5cm and 1.8cm]
	\node[block] (prm) {
		\begin{tabular}{c}
		\textbf{PRMSolver}\\
		generate nodes\\
		store graph
		\end{tabular}
	};
	\node [block, right=of prm] (config) {
		\begin{tabular}{c}
		\textbf{Configpace}\\
		indicator fct. \\
		interface
		\end{tabular}
	};
	\node [block, right=of config] (robo) {
		\begin{tabular}{c}
		\textbf{RobotConfigspace}\\
		robot kinematics\\
		collision detection
		\end{tabular}
	};
	\node [block, right=of robo] (gpu) {\textbf{GPU}};

	\path[-stealth,thick, shift left=1.0ex]	(prm) edge (config) ;
	\path[-stealth,thick, shift left=1.0ex]	(config) edge  (prm) ;
	\path[-stealth,thick, shift left=1.0ex]	(config) edge (robo);
	\path[-stealth,thick, shift left=1.0ex]	(robo) edge  (config);
	\path[-,thick]					(robo) edge (gpu);
\end{tikzpicture}
\end{center}
%\ \\

@@

To calculate the intersection between two convex polytopes, the Chung-Wang algorithm only needs their vertices 
and the data which vertices are connected by an edge. 
For this the adjacency matrix of the edge graph is stored compactly in the crs format.
\begin{lstlisting}
struct polytope4{
    float4* vertices;	//length n
    int n;
    //!edges in crs format
    int* dsp;	//length n
    int* cnt;	//length n
    int* dest;	//length m
    int m;
};
\end{lstlisting}
It means, that for $i=0,...,n-1$ the vertices $dest[dsp[i]],...,dest[dsp[i]+cnt[i]-1]$ are the ones connected with vertex $i$ by an edge.
The polytopes used in this projects are given by the user as convex hull of a list of vertices. However, to determine the edge graph only from the vertices is a nontrivial task, which is done by Matlab here using provided high level functions.

@@

@@

@@

here overview over the components of the programm components:

- description of the different versions:

- mpi

- every process has its own GPU

@@

In the first approach, that was evaluated, every process generates independantly from each other new nodes.
Then the possible neighbors are determined and the connections, which have to be tested.
The indicator function is called and returnes the edges.
After this the graph must be synchronized on every processes, which leads to an alltoall communication.
As in this graph building does not need the edge data, it is sufficient to exchange only the nodes in this phase.

@@

\subsubsection*{Version 2}
The second approach

	- every process generates all nodes (same by exchange of random seed)

	- indicator function splitted over processes, synchronous

@@

	- same as version 2 with some pipeline optimizations

version 4:

	- same as version 3 with asynchronous gpu

version 5:

	- each process has 2 workers which are generating their nodes an are using 1 gpu

	- one worker uses the kernel time  of the other to do its work

	- should give use the gpu computation time efficiently

Stateflow for each PRMSolver instance:

@@

- difficult to test because
	- not deterministic
	- dependant on density of nodes: how many neighbours have to be tested per connection

>>>>

The abstract problem is to find a path on a d-dimensional space, which avoids intersecting obstacles. The obstacles are  given by an indicator function
\begin{equation*}
	I: \Omega \rightarrow \{0, 1\}
\end{equation*}
on the domain \( \Omega = [ q_{min,1}, q_{max,1} ] \times ... \times  [ q_{min,d}, q_{max,d} ] \subset \R^d  \) and the goal is to find a continuous way
\begin{equation*}
	\Gamma: [0,1] \rightarrow \Omega
\end{equation*}
between a given start and endpoint, \(\Gamma(0)=q_b,\ \Gamma(1)=q_e \), 
such that
\begin{equation*}
	I(\Gamma(s))=0 \text{\ f.a.\ } s \in [0,1].
\end{equation*}
The probabilistic roadmap (PRM) approach grows a graph \(G=(V,E)\) form \(q_s\) and \(q_e\) by randomly sampling nodes \(v\in V\subset \Omega\).  Two nodes \(v,w \in V\) are connected, if they are near enough each other and the linear connection in between is free from obstacles. The algorithm terminates, when there is a path from \(q_s\) to \(q_e\) on G.
\begin{defi}
For two points \(q_1\) and \(q_2\) we define the connection with stepsize \(h>0\) as
	\begin{equation*}
		[q_1,q_2]_h \defgl \{ q = \lambda q_1+(1-\lambda) q_2\ |\ \lambda \in [0,1], \norm{q-q_1} \in \N_0 h \}.
	\end{equation*}
	We say, \(q_1\) and \(q_2\) are connected (with stepsize \(h>0\)) on the map \(\Omega\), if
	\(I(q)=0\) for all \(q \in [q_1, q_2]_h\).
\end{defi}

\begin{algorithm} 
	\caption{Probabilistic Roadmap in general}
	\KwData{ \(q_s, q_e\in \Omega, h>0, d_0>0\)}
	\KwResult{ \(q_s=q_1, ... ,q_n=q_e\in \Omega\), 
		such that all \(q_i\) and \(q_{i+1}\) are connected with stepsize \(h\)}
	Initialize graph \(G=(V,E)\) with \(V=\{q_s, q_e\}\) \\
	If \(q_s\) and \(q_e\) are connected, \(E \leftarrow\{ q_s, q_e\}\) \\
	\While{ there exists no path in G between \(q_s\) and \(q_e\)}{
		Sample \(q\in \Omega\) from a probability density  \(p(\ \cdot\ , G)\), resample until \(I(q)=0\) \\
		\For{ all nodes \(\tilde{q} \in V\), with \( \norm{q- \tilde q} \le d_0 \) }{
			\If { $q$ and $\tilde q$ connected with stepsize h} {
				\(E \leftarrow \{q, \tilde{q}\}\)
			}
		}
		\(V \leftarrow{q}\), if at least one edge was inserted \\
	}
	Determine the shortest path \(q_1, ... , q_n\in V\) on \(G\) from \(q_s\) to \(q_e\).
\end{algorithm}

@@

\begin{center} 
\begin{tikzpicture}[auto]%node distance=0.5cm and 1.8cm]
	\node[block] (prm) {
		\begin{tabular}{c}
		\textbf{PRMSolver}\\
		generate nodes\\
		store graph
		\end{tabular}
	};
	\node [block, right=of prm] (config) {
		\begin{tabular}{c}
		\textbf{Configpace}\\
		indicator fct. \\
		interface
		\end{tabular}
	};
	\node [block, right=of config] (robo) {
		\begin{tabular}{c}
		\textbf{RobotConfigspace}\\
		robot kinematics\\
		collision detection
		\end{tabular}
	};
	\node [block, right=of robo] (gpu) {\textbf{GPU}};

	\path[-stealth,thick, shift left=1.0ex]	(prm) edge (config) ;
	\path[-stealth,thick, shift left=1.0ex]	(config) edge  (prm) ;
	\path[-stealth,thick, shift left=1.0ex]	(config) edge (robo);
	\path[-stealth,thick, shift left=1.0ex]	(robo) edge  (config);
	\path[-,thick]					(robo) edge (gpu);
\end{tikzpicture}
\end{center}

@@

To calculate the intersection between two convex polytopes, the Chung-Wang algorithm only needs their vertices 
and the data which vertices are connected by an edge. 
For this the adjacency matrix of the edge graph is stored compactly in the crs format.
\begin{lstlisting}
struct polytope4{
    float4* vertices;	//length n
    int n;
    //edges
    int* dsp;	//length n
    int* cnt;	//length n
    int* dest;	//length m
    int m;
};
\end{lstlisting}
It means, that for $i=0,...,n-1$ the vertices $dest[dsp[i]],...,dest[dsp[i]+cnt[i]-1]$ are the ones connected with vertex $i$ by an edge.
The polytopes used in this projects are given by the user as convex hull of a list of vertices. However, to determine the edge graph only from the vertices is a nontrivial task, which is done by Matlab here using provided high level functions.

@@

- introduction

- dijkstra at the end makes almost no contribution to the whole time

	- can be parallelized, but this is again complicated

- focus on building the graph

@@

- describe

- compare different models...?

@@

The parallelization is done with MPI. All processes store their own versions of the graph, which are hold up to date between each other.
Furthermore, it is assumed that every process has its own GPU.
The work of the single MPI processes could be parallelized further with OpenMP, but this has not been done here.
Instead only different approaches to distribute the work on several GPUs have been investigated.

@@

In the first approach, that was evaluated, every process generates new nodes independantly from each other.
Then the possible neighbours are determined and the connections, which have to be tested.
The indicator function is called and returnes the edges.
After this the graph must be synchronized on every processes, which leads to an alltoall communication.
As in this graph building does not need the edge data, it is sufficient to exchange only the nodes in this phase.

@@

\subsubsection*{Version 2}
The second approach uses the fact, that by synchronizing the random seed, 
different processes can produce the same sequence of random numbers independently from each other.
By using this, every process can generate all new nodes and determine the connection data. 
Then the indicator function computation is splitted over their configuration space instances and the returned data shared afterwards with each other.
With this data, all graphs can be updated and it is guaranteed, that they are the same on each process.
The purpose of this version compared to the first one is, that less data has to be sent. 
Furthermore, in the first version it can occur, that the different generated nodes can result in very different counts of neighbours and therefore different loads on the GPUs.
Here, the distribution of the computation amount over the different GPUs can be chosen almost freely.

@@

The third version does a further optimization by using the asynchronous version of the indicator function.
Here, every process runs two workers overlapped. Here, in the time, when the first worker has generated his indicator function request and launched the kernel and is waiting until it has finished, the second worker updates the graph and sends his new request, and the other way round.
With this it is possible, that the computation time of the GPU is fully used, and overlapped with the communication.

@@

- difficult to test because

	- not deterministic

	- dependant on density of nodes: how many neighbours have to be tested per connection
